---
title: "Forecast on climate in New Dehli"
author: "Marcus Bergdahl"
date: "16 oktober 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load libraries, fig.cap = "", include = FALSE}
source(paste(getwd(), "/Scripts/load_packages.R", sep = ""))
source(paste(getwd(), "/Scripts/load_data.R", sep = ""))
```

## Background

- I wanted to re-use the work I did in my master thesis on new data.
- In the thesis I wrote my own functions for doing Bayesian vector autoregression which I applied on data from an A/B-test. The goal then was to early try to see the long term effects.
- Now I will use data on the climate in New Dehli which I downloaded from Kaggle.
- I have changed some things on my original code so I now also include seasonal effects and trend.
- First I will forecast based on the existing train-dataset and compare my predictions against the existing test-dataset. Then I will use less data in the train-dataset which I move to the test-dataset to see the performance for longer forecasts.

## First glimpse of the data

```{r glimpse of the data, out.width = "100%", out.height = "100%"}

data_train %>%
  gather(key, value, -date) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
    geom_line(aes(color = key), size = 1) +
    theme_minimal() +
    facet_grid(key ~ ., scales = "free_y")

```


## Removing outliers

Removing 9 dates where "meanpressure" < 950 or "meanpressure" > 1050

```{r remoiving outlier, include=FALSE}

data_train %>%
  filter(meanpressure < 950 | meanpressure > 1050) %>% 
  summarise(filtered_out_obs = n()) %>%
  as.data.frame()

```

```{r glimpse of the data after removing of outliers, out.width = "100%", out.height= "80%"}

data_train %>%
  filter(meanpressure > 950 & meanpressure < 1050) %>% 
  gather(key, value, -date) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
    geom_line(aes(color = key), size = 1) +
    theme_minimal() +
    facet_grid(key ~ ., scales = "free_y")

```

## Monthly seasonal effects

Average values for each months shows all four feature has seasonal effects 

```{r seasonal effects, out.width = "100%", out.height = "80%"}
data_train %>%
        filter(meanpressure > 950 & meanpressure < 1050) %>% 
        mutate(month_num = month(date)) %>%
        group_by(month_num) %>%
        summarise(
          mean_humidity = mean(humidity),
          mean_meanpressure = mean(meanpressure),
          mean_meantemp = mean(meantemp),
          mean_wind_speed = mean(wind_speed)) %>%
        gather(key, value, -month_num) %>%
  
        ggplot(mapping = aes(x = month_num, y = value)) + 
        geom_line(aes(color = key), stat = "identity") +
        theme_minimal() +
        facet_wrap(key ~ ., scales = "free_y", nrow=2)
```


## Features correlates with each other and have little trend effect

```{r correlation}

corrs <- data_train %>% 
          mutate(trend = 1:n()) %>%
          filter(meanpressure > 950 & meanpressure < 1050) %>% 
          select_if(is.numeric) %>% 
          filter_all(all_vars(!is.na(.))) %>% 
          cor(method = "pearson")

corrplot(as.matrix(corrs))

```


## Notes after exploration

- Feature "meanpressure" has 9 outliers which we removes
- There is a seasonality effect
- The features correlates with each other which is good for when building a Bayesian vector autoregression-model
- "temp" and "humidity" has a small correlation with time

```{r setup data, include = FALSE}

source(paste(getwd(), "/Scripts/month_dummies_function.R", sep = ""))
source(paste(getwd(), "/Scripts/data_setup.R", sep = ""))

index_ok <- data_train$meanpressure > 950 & data_train$meanpressure < 1050
setup_result <- setup_data(matrix = as.matrix(data_train[index_ok,2:5]), 
                            p = 1,
                            date = data_train[index_ok,1] %>% pull())

```

```{r BVAR, include = FALSE}

source(paste(getwd(), "/Scripts/bvar_function.R", sep = ""))

days_forecast <- 113
gibbs_iterations <- 10

BVAR_forecast_result <- BVAR_forecast(H=days_forecast,
                                        R=gibbs_iterations,
                                        Y=setup_result$Y,
                                        Z=setup_result$Z,
                                        z=setup_result$z,
                                        p=setup_result$p)

forecast <- cbind(data_test[,1],
              apply(BVAR_forecast_result[[1]], 2, median),
              apply(BVAR_forecast_result[[2]], 2, median),
              apply(BVAR_forecast_result[[3]], 2, median),
              apply(BVAR_forecast_result[[4]], 2, median))

colnames(forecast) <- c("date",colnames(data_test[,2:5]))

```


## It's time to forecast!

- Vector auoregression (VAR) uses all the metrics past values when forcasting.
- We will use a Bayesian approach with non-informative priors which lower the risk of overfitting
- Bayesian vector autoregression (BVAR) use gipps sampling which generates samples from a posterior multivariate distribution. For each prediction we will estimate the dates posterior distribution with 1000 samples.


## Forecast 

```{r forecast plot, out.width = "100%"}


rbind(data_train %>%
        filter(meanpressure > 950 & meanpressure < 1050) %>% 
        gather(key, value, -date) %>%
        mutate(type = "data"),
  
      forecast %>%
        gather(key, value, -date) %>%
            mutate(type = "forecast")) %>%
  
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = type), size = 1) +
  theme_minimal() +
  facet_grid(key ~ ., scale = "free_y")


```

## Estimation with its 95% confidence intervalls compared to real data

```{r confidence intervall, out.width = "100%"}


m <- 4
# estimation
pred <- list()
for(i in 1:days_forecast){
  pred[[i]] <- matrix(0, ncol=3, nrow=days_forecast)
  colnames(pred[[i]]) <- c('Median', 'Lower CI','Upper CI')  
}

for(j in 1:m){
  for(i in 1:days_forecast){
    pred[[j]][i,1] <- median(BVAR_forecast_result[[j]][,i])
    pred[[j]][i,2] <- sort(BVAR_forecast_result[[j]][,i])[1+round(gibbs_iterations*0.025)]
    pred[[j]][i,3] <- sort(BVAR_forecast_result[[j]][,i])[round(gibbs_iterations*0.975)]
  }
}



p1 <- rbind(
  data_test %>%
    dplyr::select(date, humidity) %>%
    gather(key, value, -date),
  
  cbind(date = data_test[,1],
                pred[[2]]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("humidity", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "humidity", y = "humidity") +
  theme(legend.title=element_blank())

p2 <- rbind(
  data_test %>%
    dplyr::select(date, meanpressure) %>%
    gather(key, value, -date),
  
  cbind(date = data_test[,1],
                pred[[4]]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("meanpressure", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "meanpressure", y = "meanpressure") +
  theme(legend.title=element_blank())

p3 <- rbind(
  data_test %>%
    dplyr::select(date, meantemp) %>%
    gather(key, value, -date),
  
  cbind(date = data_test[,1],
                pred[[1]]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("meantemp", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal()+
  labs(title = "meantemp", y = "meantemp") +
  theme(legend.title=element_blank())

p4 <- rbind(
  data_test %>%
    dplyr::select(date, wind_speed) %>%
    gather(key, value, -date),
  
  cbind(date = data_test[,1],
                pred[[3]]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("wind_speed", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "wind_speed", y = "wind_speed") +
  theme(legend.title=element_blank())

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE)
# blir fel med legends

```

## Forecasting a longer period

```{r}

cut_date <- "2015-12-31"
data_train_2 <- data_train %>% filter(date <= cut_date)
data_test_2 <- rbind(data_train %>% filter(date > cut_date), data_test)

index_train_ok <- data_train_2$meanpressure > 950 & data_train_2$meanpressure < 1050
index_test_ok <- data_test_2$meanpressure > 950 & data_test_2$meanpressure < 1050
setup_result_2 <- setup_data(matrix = as.matrix(data_train_2[index_train_ok,2:5]), 
                            p = 1,
                            date = data_train_2[index_train_ok,1] %>% pull())

days_forecast_2 <- 480
gibbs_iterations_2 <- 10

BVAR_forecast_result_2 <- BVAR_forecast(H=days_forecast_2,
                                        R=gibbs_iterations_2,
                                        Y=setup_result_2$Y,
                                        Z=setup_result_2$Z,
                                        z=setup_result_2$z,
                                        p=setup_result_2$p)

forecast_2 <- cbind(data_test_2[,1],
              apply(BVAR_forecast_result_2[[1]], 2, median),
              apply(BVAR_forecast_result_2[[2]], 2, median),
              apply(BVAR_forecast_result_2[[3]], 2, median),
              apply(BVAR_forecast_result_2[[4]], 2, median))

colnames(forecast_2) <- c("date",colnames(data_test_2[,2:5]))

```


I have now moved all the data from 2016 to the test-dataset because I wanted to test forecasting a longer period

```{r, out.width = "100%"}

rbind(data_train_2 %>%
        filter(meanpressure > 950 & meanpressure < 1050) %>% 
        gather(key, value, -date) %>%
        mutate(type = "data"),
  
      forecast_2 %>%
        gather(key, value, -date) %>%
            mutate(type = "forecast")) %>%
  
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = type), size = 1) +
  theme_minimal() +
  facet_grid(key ~ ., scale = "free_y")
```

## Estimation with its 95% confidence intervalls compared to real data

```{r confidence intervall - long forecast, out.width = "100%"}


m <- 4
# estimation
pred_2 <- list()
for(i in 1:days_forecast_2){
  pred_2[[i]] <- matrix(0, ncol=3, nrow=days_forecast_2)
  colnames(pred_2[[i]]) <- c('Median', 'Lower CI','Upper CI')  
}

for(j in 1:m){
  for(i in 1:days_forecast_2){
    pred_2[[j]][i,1] <- median(BVAR_forecast_result_2[[j]][,i])
    pred_2[[j]][i,2] <- sort(BVAR_forecast_result_2[[j]][,i])[1+round(gibbs_iterations_2*0.025)]
    pred_2[[j]][i,3] <- sort(BVAR_forecast_result_2[[j]][,i])[round(gibbs_iterations_2*0.975)]
  }
}



p1 <- rbind(
  data_test_2 %>%
    filter(meanpressure > 950 & meanpressure < 1050) %>%
    dplyr::select(date, humidity) %>%
    gather(key, value, -date),
  
  cbind(date = data_test_2[index_test_ok,1],
                pred_2[[2]][index_test_ok,]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("humidity", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "humidity", y = "humidity") +
  theme(legend.title=element_blank())

p2 <- rbind(
  data_test_2 %>%
    filter(meanpressure > 950 & meanpressure < 1050) %>%
    dplyr::select(date, meanpressure) %>%
    gather(key, value, -date),
  
  cbind(date = data_test_2[index_test_ok,1],
                pred_2[[4]][index_test_ok,]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("meanpressure", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "meanpressure", y = "meanpressure") +
  theme(legend.title=element_blank())

p3 <- rbind(
  data_test_2 %>%
    filter(meanpressure > 950 & meanpressure < 1050) %>%
    dplyr::select(date, meantemp) %>%
    gather(key, value, -date),
  
  cbind(date = data_test_2[index_test_ok,1],
                pred_2[[1]][index_test_ok,]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("meantemp", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal()+
  labs(title = "meantemp", y = "meantemp") +
  theme(legend.title=element_blank())

p4 <- rbind(
  data_test_2 %>%
    filter(meanpressure > 950 & meanpressure < 1050) %>%
    dplyr::select(date, wind_speed) %>%
    gather(key, value, -date),
  
  cbind(date = data_test_2[index_test_ok,1],
                pred_2[[3]][index_test_ok,]) %>%
    gather(key, value, -date)) %>%
  mutate(key = factor(key, levels = c("wind_speed", 'Lower CI', 'Median','Upper CI'))) %>%
  ggplot(mapping = aes(x = date, y = value)) + 
  geom_line(aes(color = key), size = 1) +
  theme_minimal() +
  labs(title = "wind_speed", y = "wind_speed") +
  theme(legend.title=element_blank())

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
# blir fel med legends

```

## Summary

- Perhaps this dataset wasn't the most challenging and you could probably solved this with a much easier method. But the purpose with this project was to play around with the code I used in my master thesis just for fun.
- If we tried some other methods we could compare the methods against each other by evaluating some metrics like for example mean square error.